Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-06-23 papers/GPUScA-2011/programming-model.tex
Changes on Thursday.

In this section we present the Fortran subset employed in this paper.  This
subsetting language will allow scientific programmers to stay within the Fortran
language and yet have direct access to GPU hardware.  We first examine how this
programming model relates to work in other languages.

@@

A number of previous efforts have exploited data-parallel programming
at the language level to utilize novel architectures, particularly in
previous decades during the reign of vector and massively parallel
computers in the high performance computing world.  The origin of the
array syntax that was adopted in Fortran 90 can be found in the APL
language.  Fortran 90 differed from previous % \cite{iverson79apl}
extensions of Fortran in that parallelism within whole-array
operations was expressed at the expression level instead of via
parallelism within explicit DO-loops (such as within IVTRAN for the
Illiac IV).

The High Performance Fortran (HPF) extension of Fortran was
proposed to add features to the language that would enhance the
ability of compilers to emit fast parallel code for distributed and
shared memory parallel computers\cite{koelbel94hpf}.  One of the
notable additions to the language in HPF was syntax to specify the
distribution of data structures amongst a set of parallel processors.
HPF also introduced an alternative looping construct to the
traditional DO-loop called {\tt FORALL} that was better suited for
parallel compilation.  An additional keyword, {\tt INDEPENDENT}, was
added to allow the programmer to indicate when the order of execution
of the program (such as a sequence of loop iterations) can be flexible
in order to allow parallel execution.  Interestingly, the parallelism
features introduced in HPF did not exploit the new array features
introduced in 1990 in any significant way, relying instead on explicit
loop-based parallelism.  This was likely in order to support parallel
programming that wasn't easily mapped onto a pure data-parallel model.

In some instances though, a purely data-parallel model is appropriate
for part or all of the major computations within a program.  One of
the systems where programmers relied heavily on higher level
operations instead of explicit looping constructs was the Thinking
Machines Connection Machine 5 (CM-5).  A common programming pattern
used on the CM-5 that we exploit in this paper was to write
whole-array operations from a global perspective in which computations
are expressed in terms of operations over the entire array instead of
a single local index.  The use of the array shift intrinsic functions
(like {\tt CSHIFT}) were used to build computations in which arrays
were combined by shifting the entire arrays instead of working based
on local offsets from single indices.  A simple 1D example is one in
which an element is replaced with the average of its own value with
that of its two direct neighbors.  Ignoring boundary indices that wrap
around, explicit indexing will result in a loop such as:

@@

An interesting line of language research that grew out of the early work
with HPF was that associated with the ZPL language work at the University
of Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar
global view of computation over arrays, but define their computations based
on the local view of indices that participate in the update of each element of
an array.

The static analysis and source-to-source transformations used are very
basic and only require the programmer to use a language subset that
employs a data-parallel programming model.  In particular, it
encourages use of four language features introduced in Fortran 90 ---
array notation, shift functions, elemental functions, and pure
procedures.  From these language constructs, we are able to easily
transform them to a lower-level CUDA or OpenCL implementation.

%This paper uses four simple features of Fortran that form a language subset
%that does not require any language extensions and can be easily transformed
%to a lower-level implementation in either CUDA or OpenCL.

@@

Fortran 90 introduced a rich array syntax that allows programmers to
write statements in terms of whole arrays or subarrays, with
data parallel operators to compute on the arrays.
Array variables can be used in expressions
based on whole-array operations.  For example, if {\tt A}, {\tt B},
and {\tt C} are all arrays of the same rank and shape and {\tt s} is a
scalar, then the statement

%% Deleted by CER
%%This has the benefit of avoiding explicit looping in the code and maintaining a high-level style that is closer to the original mathematical specification of the problem being solved.  More importantly from a compilation perspective, this defers decisions about how to implement these whole-array operations to a compilation tool.  Sophisticated loop analysis to identify parallelism within the sequential loop code are not necessary.  When faced with a novel architecture such as modern GPUs that are ideally suited to data parallel programming models, the fit between Fortran arrays and these systems is quite clean.

%% Deleted by CER
%% A Fortran array combines a base pointer to memory with metadata that describes the array shape, stride and size. It should be noted that other languages that provide rich array data types and whole array operations may also be suitable targets for the transformations described in this paper.

@@

%% NEW - CER
%%% The use of array notation allows one to program in a data-parallel subset of Fortran. This style of programming makes use of array notation and pure and elemental functions to operate on array elements.

@@

%% NEW - CER
%%% Because elemental functions return scalar values and are free from side effects, the compiler is free to distribute the computation over any hardware processing elements available to it, such as the multiple cores and vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.

%% NEW - CER
For example, the basic array operation shown above could be refactored into
an elemental function,

@@

%Also note that the plus and times operators shown in the data-statement
%example can be seen as an elemental functions, as plus and times are defined
%in terms of scalars but can be applied to whole arrays and can return an array
%result.

@@

\subsubsection*{Shift functions}

Many array-based algorithms require the same operation to be performed
on each element of the array using the value of that element and some small
set of neighboring cells. As described above, this can often be done with
shift operations.
Fortran provides a set of shifting operators that allow programmers to
define operations based on shifted arrays.  These intrinsic operators take
an array, a dimension, and the amount by which it should be shifted (using
the sign to indicate direction).  By defining operations on entire arrays
based on a global view of them shifted relative to each other, programmers can
avoid explicit looping and potentially tricky index arithmetic.  Furthermore,
analysis of the extent of the set of shifted arrays in a given expression
allows analysis tools to determine the amount of temporary or buffer storage
necessary to hold intermediate values during whole array operations.  With
explicit loops, programmers must maintain this temporary storage manually.

%% Deleted by CER
%% Often programmers implement these operations that are local to each element within the inner-loop of a set of nested FOR- or DO-loops using offsets relative to the current array index.  An alternative to this local-view of the algorithm is to take a global view and write the algorithm in terms of the whole array.  For example, consider a 1D array in which we wish to subtract the $(i-1)$th element from the $i$th for all elements.  One way to look at this is that we are subtracting the entire array shifted by one element from itself.

@@

Unlike OpenMP, we require \emph{limited} use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:

\item {\tt \!\$OFP CONTIGUOUS}: specifies that an dummy array variable is
  contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.

\item {\tt \!\$OFP KERNEL}: specifies that a pure subroutine can be
  transformed to an OpenCL kernel.

@@

Three new procedures (in addition to the intrinsic shift function) are
defined in Fortran that are used in array-syntax operations.  Each
procedure takes a integer array halo argument that specifies the
number of ghost cells on either side of a region, for each dimension.
For example {\tt halo = [left, right, down, up]} specifies a halo for
a two-dimensional region.  These functions are:

@@

\item {\tt transfer\_halo(array, halo)}: an impure subroutine that
  exchanges halo regions between nodes using MPI or Fortran
  coarrays. Array is the memory stored on the node representing the
  node-local portion of a virtual global array. Not used in this work.

\item {\tt interior(array, halo)}: a pure function that returns a copy of
  the interior portion of the array specified by halo. Array is
  local to a node.

\item {\tt region(array, halo)}: a pure function that returns a copy of
  the portion of the array specified by halo.  Array is local to a node.

@@

It should be noted that the two functions {\tt interior} and {\tt
  region} are pure and thus can be called from within a pure kernel
procedure.  These two functions are part of the language recognized by
the compiler and though the two functions return a copy of a portion
of an array \emph{semantically}, the compiler is not forced to
actually make a copy and is free to enforce copy semantics through
other means.

@@

%% MJS - mention that these features were designed for parallelism in
%% the first place?

%% Deleted by CER
%% An example of this style of programming in Fortran is shown in
%
%{\small
%\begin{verbatim}
%  Bz = Bz &
%     + dt*(cshift(Ex,dim=2,shift=+1)-Ex)/dy &
%     - dt*(cshift(Ey,dim=1,shift=+1)-Ey)/dx
%\end{verbatim}
%}

%%This example is a solution to Maxwell's equations for the $z$ component of the magnetic field using Fortran array syntax.  Note that there are no explicit loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt Ex}, and {\tt Ey}, individually.

%Blelloch~\cite{blelloch90,rajopadhye93}.

@@

%%~\cite{simonpeytonjones}).  The heart of the solution
%to Maxwell's equations is the statement shown in
%Listing~\ref{lst:dpexample} and five similar, simple equations.

%% NEW - CER
%%% The power in this notation is that the compiler is free to distribute the computation in these expressions over any hardware processing elements available to it, such as the vector units on an Intel or AMD processor or the Synergistic Processing Units on an IBM Cell processor.  If the arrays are declared as coarrays, this includes spreading the computation over \emph{all of the nodes in   a cluster as well.}  In this case, communication occurs within the \texttt{cshift} functions, though the compiler is free to overlap communication and computation by scheduling the communication early and performing the computation on the interior of the arrays while waiting for the communication to complete.

%% NEW - CER
%Complete and very concise and elegant programs can be built with procedures
%similar to the example shown above. To aid this effort, Fortran
%supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
%\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
%functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
%functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
%\texttt{PRODUCT}, ...).  To this set we add region functions described above.

%% Deleted by CER
%% This style of programming meets the requirements we have set for a programming model for developing applications suitable for acceleration.  It allows the programmer to program at a very-high level of abstraction while providing the compiler with maximum flexibility in targeting the application for a particular hardware architecture.  The data parallel programming model simultaneously meets the seemingly conflicting goals of maintainability, portability, and performance.

Unfortunately, this style of programming has never really caught on
because when Fortran 90 was first introduced, performance of these
features was relatively poor and thus programmers shied away from
using array syntax (even recently, some are actively counseling against its
usage because of performance issues~\cite{Levesque:SC08}).  Thus the
Fortran community was caught in a classic ``chicken-and-egg''
conundrum: (1) programmers didn't use it because it was slow; and (2)
compilers vendors didn't improve it because programmers didn't use it.
A goal of this paper is to demonstrate that parallel programs written
in this style of Fortran are maintainable and can achieve good
performance on accelerator architectures.

\subsection{Current Limitations}

@@

%% Duplicated CER
%%Only elemental functions may be called from kernel functions.  These include Fortran functions that have an OpenCL analog and user-defined elemental functions.

A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
pure and all array variables must be declared as contiguous.  A kernel
procedure may not call other procedures except for limited intrinsic
functions (primarily math), user-defined elemental functions, and the
{\tt interior} and {\tt region} functions.  We plan to address
non-contiguous arrays (such as those that result from strided access)
by mapping array strides to gather/scatter-style memory accessors.

>>>>

In this section we present the Fortran subset employed in this paper.  This
subsetting language will allow scientific programmers to stay within the Fortran
language and yet have direct access to GPU hardware.  We first examine how this
programming model in Fortran relates to developments in other languages.

@@

A number of previous efforts have exploited data-parallel programming at the
language level to utilize novel architectures, particularly in previous decades
during the reign of vector and massively parallel computers in high performance
computing (HPC).  The origin of the array syntax adopted by Fortran in the 1990
standard can be found in the APL language~\cite{iverson79apl}.  These additions
to Fortran allowed parallelism to be expressed with whole-array operations at
the expression level, instead of via parallelism within explicit DO-loops (such
as within IVTRAN for the Illiac IV).

The High Performance Fortran (HPF) extension of Fortran was proposed to add
features to the language that would enhance the ability of compilers to emit
fast parallel code for distributed and shared memory parallel
computers\cite{koelbel94hpf}.  One of the notable additions to the language in
HPF was syntax to specify the distribution of data structures amongst a set of
parallel processors.  HPF also introduced an alternative looping construct to
the traditional DO-loop called {\tt FORALL} that was better suited for parallel
compilation.  An additional keyword, {\tt INDEPENDENT}, was added to allow the
programmer to indicate when the loop contained no loop-order dependencies that
allowed for parallel execution.  These constructs are similar to {\tt DO
  CONCURRENT}, an addition to Fortran in 2008.  Interestingly, the parallelism
features introduced in HPF did not exploit the new array features introduced in
1990 in any significant way, relying instead on explicit loop-based parallelism.
This restriction allowed the language to support parallel programming that
wasn't easily mapped onto a pure data-parallel model.

In some instances though, a purely data-parallel model is appropriate for part
or all of the major computations within a program.  One of the systems where
programmers relied heavily on higher level operations instead of explicit
looping constructs was the Thinking Machines Connection Machine 5 (CM-5).  A
common programming pattern used on the CM-5 (that we exploit in this paper) was to
write whole-array operations from a global perspective in which computations are
expressed in terms of operations over the entire array instead of a single local
index.  The use of the array shift intrinsic functions (like {\tt CSHIFT}) were
used to build computations in which arrays were combined by shifting the entire
arrays instead of working on local offsets based on single indices.  A simple
1D example is one in which an element is replaced with the average of its own
value and that of its two direct neighbors.  Ignoring boundary indices that
wrap around, explicit indexing will result in a loop such as:

@@

An interesting line of language research that grew out of the early work with
HPF was that associated with the ZPL language work at the University of
Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar global
view of computation over arrays, but define their computations based on regions,
which provide a local view of the set of indices that participate in the update
of each element of an array.  The Chapel parallel programming language being
developed by Cray is directly influenced by HPF and
ZPL~\cite{chamberlain_chapel}.

\subsection{Fortran Language Subset}

The static analysis and source-to-source transformations used in this work are
very basic and only require the programmer to use a language subset that employs
a data-parallel programming model.  In particular, it encourages the use of
array notation, elemental functions, and pure procedures.  From these language
constructs, we are able to easily transform Fortran procedures to a lower-level
OpenCL kernel implementation.

@@

Fortran has a rich array syntax that allows programmers to write statements in
terms of whole arrays or subarrays, with data parallel operators to compute on
the arrays.  Array variables can be used in expressions based on whole-array
operations.  For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays of the
same rank and shape and {\tt s} is a scalar, then the statement

@@

@@

For example, the basic array operation shown above could be refactored into an
elemental function,

@@

@@

@@

@@

ForOpenCL provides two new procedures that are defined in Fortran and are used
in array-syntax operations.  Each procedure takes a integer array halo argument
that specifies the number of ghost cells on either side of a region, for each
dimension.  For example {\tt halo = [left, right, down, up]} specifies a halo
for a two-dimensional region.  These functions are:

@@

\item {\tt interior(array, halo)}: a pure function that returns a copy of
  the interior portion of the array specified by halo.

\item {\tt region(array, halo)}: a pure function that returns a copy of
  the portion of the array specified by halo.

@@

It should be noted that the two functions {\tt interior} and {\tt region} are
pure and thus can be called from within a pure kernel procedure.  These two
functions are part of the language recognized by the compiler and though the two
functions return a copy of a portion of an array \emph{semantically}, the
compiler is not forced to actually make a copy and is free to enforce copy
semantics through other means.  In addition to these two new functions,
ForOpenCL provides the compiler directive, {\tt \!\$OFP KERNEL}, which specifies
that a pure subroutine can be transformed to an OpenCL kernel.

@@

@@

Unfortunately, this style of programming has never really caught on because when
array syntax was first introduced in Fortran, performance of these features was
relatively poor and thus programmers shied away from using array syntax (even
recently, some are actively counseling against its usage because of performance
issues~\cite{Levesque:SC08}).  Thus the Fortran community was caught in a
classic ``chicken-and-egg'' conundrum: (1) programmers didn't use it because it
was slow; and (2) compilers vendors didn't improve it because programmers didn't
use it.  A goal of this paper is to demonstrate that parallel programs written
in this style of Fortran are maintainable and can achieve good performance on
accelerator architectures.

\subsection{Restrictions}

@@

A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
pure and all array variables must be declared as contiguous.  A kernel procedure
may not call other procedures except for limited intrinsic functions (primarily
math), user-defined elemental functions, and the {\tt interior} and {\tt region}
functions.  Future work will address non-contiguous arrays (such as those that result
from strided access) by mapping array strides to gather/scatter-style memory
accessors.
